Accessible and Completed Datasets yet Majority are Bronze-Graded and not Updated Regularly with Missing Metadata

### 2025-05-15 paper/paper.qmd
Finished introduction and cleaned up data section

In the analysis of our paper, we looked at the following questions: What is the quality of the datasets on the Open Data Toronto portal? What are the features of different types of datasets on Open Data Toronto?

@@

## Measurement
The Information & Technology department at Open Data Toronto collected datasets on their portal using the CKAN Datastore API
[@citeCatalogue; @citeCatalogueSteps]. Open Data Toronto uses a metric called the "Data Quality Score" in order to give each dataset in their catalogue a grade indicating "bronze", "silver", or "gold", which can be seen through on the webpage for each dataset on Open Data Toronto portal website. In order to create the "Data Quality Score", they assembled the Data Quality Working Group, which consist of a diverse group of people from consumers of datasets to people who create datasets [@citeCatalogue; @citeCatalogueSteps].

@@

![Weighting](../other/images/weighting_of_dimensions.png)

For more technical details about how each dimension is calculated for a dataset and the final data quality score, it can be through the following link:[https://github.com/open-data-toronto/framework-data-quality/blob/master/data_quality_score.ipynb](https://github.com/open-data-toronto/framework-data-quality/blob/master/data_quality_score.ipynb) [@citeDQSNotebook]. However, it is worth noting that each dimension are calculated based on data obtained on the metadata information and the dataset itself. If the data quality score of a dataset has a normalized score of less than 60%, they are given the grade "Bronze", if the normalized score is between 60% and 80%, it is given the grade "Silver", and finally if the normalized score is over 80%, tit is given the grade "Gold". The score for each dataset in the portal is recalculated every week by the team [@citeCatalogue].

@@

Our variables of interest that we used in our analysis are the following: "accessibility", "completeness", "freshness", "metadata", "usability", and "grade". "Accessiblity" is a score from 0 to 1 that indicates the degree that the dataset can be access or not through the Open Data Toronto API, keywords or tags, and automated data pipelines with 1 being that it can not be accessed with the various methods noted and 0 if not at all. "Completeness" is a score from 0 to 1 that indicates how much of the data is missing with 1 being that there is no missing data and 0 being that all data fields are empty or missing. The "freshness" variable is a score from 0 to 1 that indicates how up-to-date the data where a shorter time duration between the recent refresh date and time and the previous refresh date and time before that gives a higher score. The "metadata" variable indicates how complete the following metadata fields are for a dataset from a scale of 0 to 1: Description, Limitations, Topics, Contact Email. The more metadata fields compelted, the higher the metadata score. The "usability" variable is a score from 0 to 1 indicating how easy it would be to use the dataset and this is determine by the proportion of meaningful column names or in other words column names with English words in it.

>>>>

In 2024, a story collaboration between the Investigative Journalism Foundation (IJF) and CBC published in the CBC reported that the risk of death and injuries from fires in lower-income Toronto wards was higher compared to higher-income Toronto wards [@citeCityFire]. This story as well as other stories published in the news rely on data from Open Data Toronto to bring their stories to life [@citeCityFire]. Open Data Toronto serves as a hub for all types of data related to Toronto from crime data to information about shelters across the city [@citeODTabout]. Open Data Toronto has not only been used in the news but also in civic spaces when establishing city policies [@citeODTabout]. Given the importance of the hub, it raises concerns about the quality of Open Data Toronto's data catalogue and the following question: What is the quality of the datasets like on the Open Data Toronto portal?

In this paper, we analyzed data provided by Open Data Toronto on the data quality grade of the datasets in their catalogue and their characteristics like "accessibility", "completeness", "freshness", "metadata", and "usability". Farrow analyzed the data quality scores of the different datasets on the Open Data Toronto portal and found that the metadata and freshness scores were poor overall [@citeFarrow]. However there has been no analysis done on the scores and how they compare for different graded datasets as of 2025. In our findings, all datasets in Open Data Toronto portal are accessible and a majority of them having minimial missing data and are usable. It also showed that 56% of their datasets are graded bronze with bronze-graded datasets being less likely to be updated and have completed metadata fields. These bronze-graded datasets contribute the poor metadata and freshness scores we saw. These findings can inform Open Data Toronto about specific graded datasets that should be given more attention and the specific qualities of them such as metadata completion, which impacts the quality of the data used in the media and in policymaking. These findings can also be informative for users of the portal in order to understand what goes behind the grade of datasets on the portal and what they mean.

@@

## Measurement
Open Data Toronto uses a metric called the "Data Quality Score" in order to give each dataset in their catalogue a grade indicating "bronze", "silver", or "gold", which can be seen through on the webpage for each dataset on Open Data Toronto portal website. In order to create the "Data Quality Score", they assembled the Data Quality Working Group, which consist of a diverse group of people from consumers of datasets to people who create datasets [@citeCatalogue; @citeCatalogueSteps].

@@

![Weight of the selected dimensions for data quality assessment and the metrics for each dimension](../other/images/weighting_of_dimensions.png)

For more technical details about how each dimension is calculated for a dataset and the final data quality score, it can be through the following link:[https://github.com/open-data-toronto/framework-data-quality/blob/master/data_quality_score.ipynb](https://github.com/open-data-toronto/framework-data-quality/blob/master/data_quality_score.ipynb) [@citeDQSNotebook]. However, it is worth noting that each dimension are calculated based on data obtained on the metadata information and the dataset itself. If the data quality score of a dataset has a normalized score of less than 60%, they are given the grade "Bronze", if the normalized score is between 60% and 80%, it is given the grade "Silver", and finally if the normalized score is over 80%, tit is given the grade "Gold". The Information & Technology department at Open Data Toronto collected datasets on their portal using the CKAN Datastore API in order to give each of them a data quality score and grade [@citeCatalogue; @citeCatalogueSteps]. The score for each dataset in the portal is recalculated every week by the team [@citeCatalogue].

@@

Our variables of interest that we used in our analysis are the following: "accessibility", "completeness", "freshness", "metadata", "usability", and "grade". "Accessiblity" is a score from 0 to 1 that indicates the degree that the dataset can be access or not through the Open Data Toronto API, keywords or tags, and automated data pipelines with 1 being that it can not be accessed with the various methods noted and 0 if not at all. "Completeness" is a score from 0 to 1 that indicates how much of the data is missing with 1 being that there is no missing data and 0 being that all data fields are empty or missing. The "freshness" variable is a score from 0 to 1 that indicates how up-to-date the data where a shorter time duration between the recent refresh date and time and the previous refresh date and time before that gives a higher score. The "metadata" variable indicates how complete the following metadata fields are for a dataset from a scale of 0 to 1: Description, Limitations, Topics, Contact Email. The more metadata fields compelted, the higher the metadata score. The "usability" variable is a score from 0 to 1 indicating how easy it would be to use the dataset and this is determine by the proportion of meaningful column names or in other words column names with English words in it. However a limitation of the "usability" variables is that it does not consider datasets that are in different languages other than English.
